Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-07-22 threshold/draft.Rnw
Added notes on general matrices (scaling and centering)

\begin{algorithm}\label{prune}
Input: data matrix $A\in \mathcal{R}^{m\times n}$, correlation threshold $t$,
truncated SVD rank $p<<k$.
\begin{enumerate}
\item Compute the truncated SVD $AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$,
where $V_{1:p} = [v_1, v_2, \ldots, v_p]$
and $\mathrm{diag}(s_{1:p})$ denotes the diagonal matrix of the first $p$ singular values.
\item Compute permutation $P$ that orders points in $v_1$.
\item Compute $\ell$, the longest run of successive points in $P v_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$.
\item Compute a set of candidate pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) : 
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
\right\},
\]
where $e_{\tilde{j}}\in\mathcal{R}^n$ consists of $1$ in position $\tilde{j}$ and
zeros otherwise.
\item If there are judged to be  ``too many'' candidate pairs, increase $p$ and go to
step 1. Continue to increase $p$ in this way until the number of candidate
pairs stops decreasing (the best this algorithm can do).
\item Recover the non-permuted candidate column vector
indices $(j,i)$ from $j=e_{\tilde{j}}^T Pq$ and $i=e_{\tilde{j}+\tilde{i}}^TPq$ for every
pair $(\tilde{j},\tilde{j} + \tilde{i})$ in step 4, where
$q\in\mathcal{R}^n = [1,2,\ldots,n]^T$.
\item Compute the full correlation coefficient for each column vector pair identified in
the previous step, applying the correlation threshold to this reduced set of values.
\end{enumerate}
\end{algorithm}
Because the singular values $\sigma_j$, $j=1,2,\ldots,n$ are non-increasing and
decrease in proportion to variability in the data, it's often reasonable to
judge whether there are ``too many'' candidate pairs (step 5) after only a 1-d
projection $\tilde{i}=1$ in step 4 of the algorithm.

@@

\section{Fast truncated SVD for general matrices}\label{irlba}

We have so far assumed that the mean of each column vector of the $m\times n$
matrix $A$ is zero and the columns are scaled to have unit norm, but we really
want a method that works with general matrices $A$. We've also implicitly
assumed that we can efficiently compute a relatively low-rank truncated SVD of
potentially large  matrices required by step 1 of Algorithm \ref{prune}.
Fortunately, both needs are met by one algorithm.

@@

We relax the column mean and scale assumptions without much additional
computational or storage overhead as follows.  Let
$x=[x_1,x_2,\ldots,x_n]\in\mathcal{R}^n$ represent the vector of column means
of the matrix $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$ be the vector
of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$ a diagonal
matrix with diagonal entries $1/\sqrt{w_i^2 - m x_i^2}$.  Let
$e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ex^T)W$ is a centered
matrix with zero column means, and scaled to have unit column norms, and
\[
\mathrm{cor}(A) = W^T (A-ex^T)^T (A-ex^T) W.
\]
IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
products. The main idea behind efficient application of IRLBA to correlation
problems replaces matrix-vector products of the form $Az$ with $AWz - ex^TWz$ in
the iterations, implicitly working with a scaled and centered matrix without
forming it.  This comes at the cost of storing two additional length $n$ vectors and
computing two additional vector inner products per matrix-vector product.

>>>>

\begin{algorithm}\label{prune}
Input: data matrix $A\in \mathcal{R}^{m\times n}$, correlation threshold $t$,
truncated SVD rank $p<<k$.
\begin{enumerate}
\item Compute the truncated SVD $AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$,
where $V_{1:p} = [v_1, v_2, \ldots, v_p]$
and $\mathrm{diag}(s_{1:p})$ denotes the diagonal matrix of the first $p$ singular values.
\item Compute permutation $P$ that orders points in $v_1$.
\item Compute $\ell$, the longest run of successive points in $P v_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$.
\item Compute a set of candidate pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) : 
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
\right\},
\]
where $e_{\tilde{j}}\in\mathcal{R}^n$ consists of $1$ in position $\tilde{j}$ and
zeros otherwise.
\item If there are judged to be  ``too many'' candidate pairs, increase $p$ and go to
step 1. Continue to increase $p$ in this way until the number of candidate
pairs is sufficiently small or stops decreasing (the best this algorithm can do).
\item Recover the non-permuted candidate column vector
indices $(j,i)$ from $j=e_{\tilde{j}}^T Pq$ and $i=e_{\tilde{j}+\tilde{i}}^TPq$ for every
pair $(\tilde{j},\tilde{j} + \tilde{i})$ in step 4, where
$q\in\mathcal{R}^n = [1,2,\ldots,n]^T$.
\item Compute the full correlation coefficient for each column vector pair identified in
the previous step, applying the correlation threshold to this reduced set of values.
\end{enumerate}
\end{algorithm}
Because the singular values $\sigma_j$, $j=1,2,\ldots,n$ are non-increasing and
decrease in proportion to variability in the data, it's often reasonable to
judge whether there are ``too many'' candidate pairs (step 5) after only a 1-d
projection $\tilde{i}=1$ in step 4 of the algorithm.

@@

\section{General matrices and fast truncated SVD}\label{irlba}

We have so far assumed that the mean of each column vector of the $m\times n$
matrix $A$ is zero and the columns are scaled to have unit norm, but we really
want a method that works with general matrices $A$.  We also want a way to
efficiently compute a relatively low-rank truncated SVD of potentially large
matrices required by step 1 of Algorithm \ref{prune}.  Fortunately, both
desires are met by one algorithm.

@@

We relax the column mean and scale assumptions without introducing much additional
computational or storage overhead as follows.  Let
$x=[x_1,x_2,\ldots,x_n]\in\mathcal{R}^n$ represent the vector of column means
of the matrix $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$ be the vector
of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$ a diagonal
matrix with diagonal entries $1/\sqrt{w_i^2 - m x_i^2}$.  Let
$e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ex^T)W$ is a centered
matrix with zero column means, and scaled to have unit column norms, and
\[
\mathrm{cor}(A) = W^T (A-ex^T)^T (A-ex^T) W.
\]
IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
products. The main idea behind efficient application of IRLBA to correlation
problems replaces matrix-vector products of the form $Az$ with $AWz - ex^TWz$ in
the iterations, implicitly working with a scaled and centered matrix without
forming it.  This comes at the cost of storing two additional length $n$ vectors and
computing two additional vector inner products per matrix-vector product.
